

## Visualize higher dimensional associations


```{r}
# Load necessary library
library(ggplot2)
```

###  ----- 1. Two Categorical Variables -----
Example data for two categorical variables

```{r}
data_cat <- data.frame(
  Gender = c("Male", "Female", "Male", "Female", "Male", "Female"),
  Preference = c("A", "B", "A", "A", "B", "B")
)
```


```{r}
# Bar plot
p1 <- ggplot(data_cat, aes(x = Gender, fill = Preference)) +
  geom_bar(position = "dodge") +
  labs(title = "Preference by Gender", x = "Gender", y = "Count") +
  theme_minimal()

p1
```

### ----- 2. Two Numerical Variables -----

 Example data for two numerical variables
```{r}

data_num <- data.frame(
  Age = c(21, 25, 30, 35, 40, 45),
  Salary = c(3000, 3200, 4000, 5000, 6000, 6500)
)
```


```{r}
# Scatter plot
p2 <- ggplot(data_num, aes(x = Age, y = Salary)) +
  geom_point(color = "blue", size = 3) +
  labs(title = "Scatter Plot: Age vs Salary", x = "Age", y = "Salary") +
  theme_minimal()
p2
```

### ----- 3. Categorical and Numerical Variable -----
Example data for a categorical and a numerical variable

```{r}
data_cat_num <- data.frame(
  Group = c("A", "A", "B", "B", "C", "C"),
  Score = c(80, 85, 78, 82, 90, 95)
)

# Box plot
p3 <- ggplot(data_cat_num, aes(x = Group, y = Score, fill = Group)) +
  geom_boxplot() +
  labs(title = "Box Plot: Score by Group", x = "Group", y = "Score") +
  theme_minimal()

p3
```
In the histogram and boxplot exercise, you added an extra layer (graph) to a similar plot that allowed us to overlay the exact data points. Inlude it in this example

```{r}

#p3_new <- ggplot(data_cat_num, aes(x = Group, y = Score, fill = Group)) +
#  geom_boxplot() +
#  # ------------- #
#  labs(title = "Box Plot: Score by Group", x = "Group", y = "Score") +
#  theme_minimal()

#p3_new
```



### ----- 4. Three Variables (Categorical + Numerical + Numerical) -----

Example data for three variables (by overlying an exta aesthetic - colour - we can now acknowledge three different variables)
```{r}
data_three <- data.frame(
  Age = c(21, 25, 30, 35, 40, 45),
  Salary = c(3000, 3200, 4000, 5000, 6000, 6500),
  Gender = c("Male", "Female", "Male", "Female", "Male", "Female")
)

p4 <- ggplot(data_three, aes(x = Age, y = Salary, color = Gender)) +
  geom_point(size = 3) +
  labs(title = "Scatter Plot: Age vs Salary by Gender", x = "Age", y = "Salary") +
  theme_minimal()

p4
```

### ----- 5. Three Variables (Two Numerical + One Categorical) -----
Example data for faceted scatter plot 


```{r}
data_facet <- data.frame(
  Age = c(21, 25, 30, 35, 40, 45, 21, 25, 30, 35, 40, 45),
  Salary = c(3000, 3200, 4000, 5000, 6000, 6500, 3100, 3300, 4100, 5100, 6100, 6600),
  Gender = rep(c("Male", "Female"), each = 6)
)

# Faceted scatter plot
p5 <- ggplot(data_facet, aes(x = Age, y = Salary)) +
  geom_point(size = 3, color = "blue") +
  facet_wrap(~ Gender) +
  labs(title = "Faceted Scatter Plot: Age vs Salary by Gender", x = "Age", y = "Salary") +
  theme_minimal()

p5
```



