## (2) Correlation - R

Again, this practical is also in R. This is an example of the great amount of resources available in open-source programming languages (will also be able to find similar things in Python), where new packages are developed and made available constantly. As key insight, explore, probably what you need is already out there for you to use!

------------------------------------------------------------------------

For you to see the breadth of possibilities out there, here is an example of different ways in which to describe and learn about the correlation matrix of your dataset.

```{r, message=FALSE}


# Install required libraries (if not already installed)

#install.packages(c("ggplot2", "corrplot", "ggcorrplot", 
#"PerformanceAnalytics", "GGally", "psych", "corrr"))

# Load necessary libraries
library(ggplot2)
library(corrplot)
library(ggcorrplot)
library(PerformanceAnalytics)
library(GGally)
library(psych)
library(corrr)
```

To assess the association between two variables, you can produce a scatter plot like below:

```{r}
set.seed(123)
x <- rnorm(100)
y <- x + rnorm(100, sd = 0.5)

ggplot(data = data.frame(x, y), aes(x = x, y = y)) +
  geom_point(color = "blue", size = 2) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(title = "Correlation Plot with ggplot2",
       x = "X values", y = "Y values") +
  theme_minimal()
```

Now we are going to produce the correlation matrix, where in a go, you get insight of all the association happening in your dataset.For this we are going to use the mtcars data (A built-in dataset that contains measurements on 11 different attributes for 32 different cars). More specifically, the chosen variables are:

-   mpg (Miles per Gallon): Represents the fuel efficiency of the car. Higher values, better efficiency.

-   disp (Displacement): Represents the engine displacement, which is the total volume of all the cylinders in the engine.

-   hp (Horsepower): Represents the power output of the car's engine.

-   wt (Weight): Represents the weight of the car.

```{r}
data <- mtcars[, c("mpg", "disp", "hp", "wt")]
```

Can you calculate the correlation between them? Try out at least a pair.Remember there are different ways in which to calculate correlation <https://www.sthda.com/english/wiki/correlation-test-between-two-variables-in-r> and you can specify the type in your code `cor(x, y, method = c("pearson", "kendall", "spearman"))`

```{r}

```

Now that you know what to expect, lets proceed with the correlation matrices (same thing as above, but all in a go!):

### -----corrplot Correlation Matrix -----

```{r}
cor_matrix <- cor(data)
corrplot(cor_matrix, method = "circle", type = "upper", 
         title = "Correlation Matrix", addCoef.col = "black")
```

### -----corrplot Correlation Matrix -----

```{r}
ggcorrplot(cor_matrix, 
           method = "square", 
           type = "lower", 
           lab = TRUE, 
           title = "mtcars", 
           lab_size = 3) +
  theme_minimal()
```

### -----PerformanceAnalytics Pairwise Correlation Plot -----

```{r}
chart.Correlation(data, histogram = TRUE, pch = 19)
```

###-----GGally Pairwise Correlation Plot -----

```{r}
ggpairs(data, 
        title = "Pairwise Correlation Plot with GGally",
        lower = list(continuous = wrap("smooth", method = "lm")),
        upper = list(continuous = wrap("cor", size = 3)))
```

###----- psych Enhanced Pairwise Correlation Plot -----

```{r}
pairs.panels(data, 
             method = "pearson", 
             hist.col = "lightblue", 
             density = TRUE, 
             ellipses = TRUE)
```

###----- Base R Heatmap -----

```{r}
# ----- 8. Base R Heatmap -----
heatmap(cor_matrix, symm = TRUE, 
        main = "Correlation Heatmap", 
        col = colorRampPalette(c("red", "white", "blue"))(20))
```

```{r}
# ----- 10. corrr Tidy Correlation -----
cor_matrix_tidy <- correlate(data)
cor_matrix_tidy %>% 
  fashion() %>% 
  print()

# Network-style correlation plot
cor_matrix_tidy %>% 
  network_plot(min_cor = 0.3)

```

What are your insights? Do they make sense? Remember:

-   mpg (Miles per Gallon): Represents the fuel efficiency of the car. Higher values, better efficiency.

-   disp (Displacement): Represents the engine displacement, which is the total volume of all the cylinders in the engine.

-   hp (Horsepower): Represents the power output of the car's engine.

-   wt (Weight): Represents the weight of the car.
